---
title: "Introducing R with behavioural inhibition data"
output:
  pdf_document: default
  html_document: default
---

## Run R code

In this file, code goes inside chunks like the one below. Click **Knit** to run all 
the chunks and make an HTML page.

```{r}
4 + 6
(4 + 6) / 2
```

You can use the assignment operator to assign values to variables. For instance, 
we can create a vector of values between 1 and 10.

```{r}
x <- 1:10
```

You can apply any mathematical operation to this vector of integers.

```{r}
x ** 2
```

```{r}
sqrt(x)
```

Like in Python, you can use `#` to comment things out. 

```{r}
# 4 + 6
# c <- 10
```

Finally, you can also run Python code inside the same environment, which is 
convenient when you are working on a project that combines both.

```{python}
x = [_ for _ in range(1, 11)]
print([i ** 2 for i in x])
```


## Objects, vectors, and functions

`c()` makes a vector. R is case sensitive: `scores` and `Scores` are different names.

```{r}
scores <- c(180, 210, 195, 205, 1)  # four SSRT values
scores
length(scores)
mean(scores)
round(x=mean(scores), digits=0)
```

Try `?mean` in the Console for help.

Use `[ ]` to select values. R starts counting at 1. A comparison returns
`TRUE` or `FALSE` for each value.

```{r}
scores[1]
scores > 200
scores[scores > 200]
```

Numbers, text, and booleans are common types. `class()` shows the type of 
a variable/value.

```{r}
class(scores)
class("Negative")
class(T)
```

`NA` means a missing value. Many functions need `na.rm = TRUE` to skip missing 
values (otherwise they return an error).

```{r}
mean(c(180, NA, 205))
mean(c(180, NA, 205), na.rm = TRUE)
```

## Data frames

A data frame has rows for people and columns for variables. `$` selects a
column; `[row, column]` selects part of a data frame.


```{r}
people <- data.frame(condition = c("Negative", "Neutral"),
                     ssrt = c(180, 210))
people
people$ssrt
people[1, 1]
```

## Load and check the study data

We are going to look at Paál et al., 2015. In this paper, they were testing a 
hypothesis that people from more deprived socio-economic backgrounds tend to be 
more impulsive. To operationalise impulsivity, they relied on the stop-signal task.
Participants have to press a key as fast as possible when a certain stimulus appears
on the screen. However, if a (stop-)signal is played, the participants need to ignore
the stimulus and not press any keys. The duration of the interval between the signal
and the stimulus appearance is varied accross participants to estimate their
inhibition period (how much time they need to stop oneself from responding). 

The experiment had 58 participants. Each person completed a stop-signal
task after a negative or neutral mood induction. They used mood as a control for 
testing the effect of the socio-economic background on inhibition. We will look
at SSRT (stop-signal reaction time) measures as a function of mood.

Install tidyverse once in the Console with `install.packages("tidyverse")`. 
Load it in each R session using the `libary()` function.

```{r message=FALSE}
library(tidyverse)

d <- read_csv("https://dfzljdn9uc3pi.cloudfront.net/2015/964/1/data_supplement.csv")
names(d)[1] <- "Participant"

dim(d)
```

Let's look at the data. 

```{r}
head(d)
```

We can also look just at column names.

```{r}
names(d)
```

Or look at snipetts of the data using the `glimpse()` function from the `dplyr` 
package.

```{r}
glimpse(d)
```

And, finally, we can look at some summary statistics for the SSRT.


```{r}
summary(d$SSRT)
```


The mood codes are `1` for Negative and `2` for Neutral. `mutate()` adds a
column.

```{r}
d$Mood_induction_condition
```


```{r}
d <- d %>%
  mutate(Condition = factor(Mood_induction_condition,
                            levels = c(1, 2), 
                            labels = c("Negative", "Neutral")))

d$Condition
```

Check the new labels with `count()`.

```{r}
d %>% 
  count(Sex)
```


`%>%` is called a pipe. It's implemented in the `magrittr` package, included 
in `tidyverse`. It passes the result on the left to the function on the right. 
That way you can combine several operations together without needing to reassign
variables or risk loosing data.

### Practice 1

Can you figure out whether it's a within- or between-participants design?
Compare the number of rows with the number of distinct participants.
`unique()` returns distinct values, and `length()` counts them.

```{r}
length(unique(d$Participant))
```


## Describe SSRT

Let's look at mean, median, standard deviation and variance of SSRT. Standard 
deviation shows spread in milliseconds. `var()` gives sample variance, 
using `n - 1`.

```{r}
d %>%
  group_by(Sex) %>% 
  mutate(mean_age = mean(Age, na.rm = TRUE))
```

`group_by()` repeats the summary for each condition. Let's look justn at the mean
and standard deviation.

```{r}
d %>%
  group_by(Condition) %>%
  summarise(mean = mean(SSRT, na.rm = TRUE),
            sd = sd(SSRT, na.rm = TRUE))
```

What can you tell about the difference between the two conditions based on these
summary statistics?

### Practice 2

Does the condition with the higher mean SSRT also have the higher median SSRT?
Use `group_by()` and `summarise()` to calculate both for each condition.

```{r}
d %>% 
  group_by(Condition) %>% 
  summarise(median_per_condition = median(SSRT, na.rm=F))
```

## Plot SSRT

For plotting, we will be using the `ggplot` package. It operates using the 
"grammar of graphics" system: each new command adds another layer to the initial
empty figure instantiated using the `ggplot()` function. 


```{r}
ggplot(d, aes(x = SSRT)) + 
  geom_histogram()
```

### Practice 3

Try making the histogram with `binwidth = 10` and then `binwidth = 40`.
How does the choice of bin width change what you notice about the SSRT values?


```{r}
ggplot(d, aes(x = SSRT)) + 
  geom_histogram(binwidth = 58)
```


### Practice 4

Can you make a separate SSRT histogram for each condition? Do the two distributions
look similar?

```{r}
ggplot(d, aes(x = SSRT, fill=Condition)) + 
  geom_histogram(binwidth = 10)
```

## Source

Based on Chapters 1–2 of Daniel Nettle, *From Questions to Knowledge: Data
Analysis for Psychology and Behavioural Science using R* (2025),
[online edition](https://bookdown.org/danielnettle2/data_analysis/). Data from
Paál, Carpenter, and Nettle (2015),
[“Childhood socioeconomic deprivation, but not current mood, is associated
with behavioural disinhibition in adults”](https://doi.org/10.7717/peerj.964).
