---
title: "Inferential statistics with behavioural inhibition data"
output:
  pdf_document: default
  html_document: default
---

## From a sample to a population

In the first practical, we looked at stop-signal reaction times (SSRTs) for the
58 participants in Paál et al. (2015). A longer SSRT means slower inhibition.
Now we want to ask what these observations might tell us about people beyond
this study. The **sample** is the participants we measured; a **population**
is the wider group about which we want to learn. A population quantity, such
as its mean SSRT, is a **parameter**. The sample mean is an **estimate** of it.

We are using the same data as last week. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)

d <- read_csv(
  "https://dfzljdn9uc3pi.cloudfront.net/2015/964/1/data_supplement.csv",
  show_col_types = FALSE
) %>%
  rename(Participant = 1) %>%
  mutate(Condition = factor(
    Mood_induction_condition,
    levels = c(2, 1),
    labels = c("Neutral", "Negative")
  ))

d %>% select(Participant, Age, Condition, SSRT) %>% slice_head(n = 6)
```

### Practice 1

The sample contains 58 participants. Name a wider population you might
reasonably want to learn about. Is `mean(d$SSRT)` a sample statistic or a
population parameter?

## Repeated samples and the central limit theorem

Imagine taking many independent samples of the same size and calculating a
mean SSRT from each. Those means form a **sampling distribution**. We cannot
recruit many new groups here, so we will simulate sampling *with replacement*
from the 58 observed SSRTs. For this simulation, their distribution stands in
for a population; it does not establish that the study represents all adults.

`slice_sample()` draws rows; `map_dbl()` repeats the calculation and stores one
mean from each sample. The seed lets everyone get the same simulation.

The central limit theorem explains why the distribution of sample means
approaches a Normal shape as sample size grows, under suitable conditions,
even when individual observations are not Normal.

```{r}
d_ssrt <- d %>% filter(!is.na(SSRT))
ssrt <- d_ssrt %>% pull(SSRT)
set.seed(2026)
means_5 <- map_dbl(1:1000, ~ d_ssrt %>%
  slice_sample(n = 5, replace = TRUE) %>%
  summarise(mean_SSRT = mean(SSRT)) %>% pull(mean_SSRT))
means_30 <- map_dbl(1:1000, ~ d_ssrt %>%
  slice_sample(n = 30, replace = TRUE) %>%
  summarise(mean_SSRT = mean(SSRT)) %>% pull(mean_SSRT))

sampling_means <- tibble(
  mean_SSRT = c(means_5, means_30),
  sample_size = factor(rep(c(5, 30), each = 1000))
)

ggplot(sampling_means, aes(x = mean_SSRT)) +
  geom_histogram(binwidth = 5, colour = "white") +
  facet_wrap(~ sample_size) +
  labs(x = "Sample mean SSRT (ms)", y = "Number of samples",
       title = "Sampling distributions for two sample sizes")
```

### Practice 2

Compare the two samples: which set of means varies more? 

```{r}
### YOUR CODE HERE ###
```

## Standard error: how much an estimate varies

The **standard deviation** (SD) describes variation among individual SSRTs.
The **standard error** (SE) describes variation among estimates from repeated
samples. For a sample mean, we estimate it from one sample using

$$SE(\bar{x}) = \frac{s}{\sqrt{n}},$$

where $s$ is the sample SD and $n$ is the number of observations. The SD of
our simulated means should be close to this formula (they need not match
exactly because we simulated only 1,000 samples).

```{r}
tibble(
  sample_size = c(5, 30),
  SD_of_simulated_means = c(sd(means_5), sd(means_30)),
  SE_from_formula = sd(ssrt) / sqrt(c(5, 30))
)
```

### Practice 3

Use the formula to predict the SE for samples of 20 and 80 people. If sample
size becomes four times as large, what happens to the SE? Try the calculation
in R before answering.

```{r}
### YOUR CODE HERE ###
```

## A confidence interval for mean SSRT

We have estimated mean SSRT last week. The lecture used $\bar{x} \pm 1.96 SE$, 
a Normal approximation. Let's compute the confidence interval for the mean SSRT.

```{r}
n <- length(ssrt)
mean_ssrt <- mean(ssrt)
se_ssrt <- sd(ssrt) / sqrt(n)

mean_ssrt + 1.96 * se_ssrt
mean_ssrt - 1.96 * se_ssrt
```

### Practice 4

We can also check the long-run meaning of a confidence interval. The following
code draws 1,000 samples of 30 from our stand-in population, calculates 
a confidence interval, and counts how many intervals include the
stand-in population mean. 

```{r}
set.seed(2024)
intervals <- map_dfr(1:1000, ~ {
  d_ssrt %>%
    slice_sample(n = 30, replace = TRUE) %>%
    summarise(mean_SSRT = mean(SSRT),
              margin = 1.96 * sd(SSRT) / sqrt(n()),
              lower = mean_SSRT - margin,
              upper = mean_SSRT + margin)
})

intervals %>%
  summarise(proportion = mean(lower <= mean(ssrt) &
                                upper >= mean(ssrt)))
```

About what proportion of intervals covered the stand-in population mean?
Would you expect exactly 0.95 every time you ran the simulation? Why?

## The simplest General Linear Model

A model with only an intercept writes each observation as

$$SSRT_i = \beta_0 + \epsilon_i.$$

$\beta_0$ is the mean SSRT we are estimating; $\epsilon_i$ is the difference
between a person's SSRT and the model's prediction. `lm()` chooses the
intercept that minimizes the **sum of squared residuals**. 

```{r}
m0 <- lm(SSRT ~ 1, data = d)
coef(m0)
confint(m0)
```

### Practice 5

Is the fitted intercept the same as `mean(ssrt)`? 

```{r}
### YOUR CODE HERE ###
```

## Adding a predictor: age and SSRT

The lecture added weight as a predictor of height. With our dataset, we can
instead ask whether mean SSRT changes with age:

$$SSRT_i = \beta_0 + \beta_1 Age_i + \epsilon_i.$$

$\beta_1$ is the expected change in SSRT, in milliseconds, for one extra year
of age. One participant has no age recorded, so we use complete age and SSRT
rows for this model. Start by excluding the participant with the missed value 
in the `Age` column. Save the new dataframe in the `d_age` variable.

```{r}
### YOUR CODE HERE ###
```


Then fit a regression with `Age` as slope and with an intercept. Save it as 
`m_age`. 

```{r} 
### YOUR CODE HERE ###
```

Now we can look at the estimated parameter values, as well at their confidence
intervals:

```{r}
summary(m_age)$coefficients
confint(m_age)
```

Let's look at the regression line relative to data:


```{r}
ggplot(d_age, aes(x = Age, y = SSRT)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age (years)", y = "SSRT (ms)")
```

The raw intercept predicts SSRT at age zero, outside the ages in this study. This
is meaningless, as you have no interest in knowing a newborn's SSRT for this
study. What you can do instead is **center** the age at the mean value. Now
the intercept estimates mean SSRT at the sample's average age, and the slope
will remain the same.

```{r}
d_age <- d_age %>% mutate(Age_centred = Age - mean(Age))
m_age_centred <- lm(SSRT ~ Age_centred, data = d_age)
summary(m_age_centred)$coefficients
```

### Practice 6

Interpret the age slope and its 95% confidence interval in words. What does
the intercept of the centered model mean? Does the association establish that
getting older *causes* slower inhibition? Explain your answer.

## Return to the mood comparison

The first practical compared SSRT descriptively across mood conditions. We
can now estimate the difference with a model. `Neutral` is the reference
category, so the coefficient for `ConditionNegative` is **Negative minus
Neutral**, in milliseconds.

```{r}
m_mood <- lm(SSRT ~ Condition, data = d)
summary(m_mood)$coefficients
confint(m_mood)
```

### Practice 7

Use `group_by()` and `summarise()` to find mean SSRT in each mood condition.
Subtract the Neutral mean from the Negative mean. Does this equal the
`ConditionNegative` coefficient?

```{r}
### YOUR CODE HERE ###
```

## A quick look at model assumptions

The model assumes approximately Normal errors. A histogram of its residuals
can reveal strong asymmetry or unusual values. It cannot check independence
or whether residual spread stays constant across fitted values.

```{r fig.height=3}
model_checks <- tibble(residual = resid(m_age))

ggplot(model_checks, aes(x = residual)) +
  geom_histogram(binwidth = 20, boundary = 0, colour = "white") +
  labs(x = "Residual SSRT (ms)", y = "Number of participants")
```

### Practice 8

Is the residual distribution roughly symmetric? Are there unusual values?

## Sources

This practical follows the topics in the class on inferential statistics and
the General Linear Model, and Chapters 2--3 of Daniel Nettle,
[*From Questions to Knowledge: Data Analysis for Psychology and Behavioural
Science using R*](https://bookdown.org/danielnettle2/data_analysis/).
The dataset is from Paál, Carpenter, and Nettle (2015),
[*Childhood socioeconomic deprivation, but not current mood, is associated
with behavioural disinhibition in adults*](https://doi.org/10.7717/peerj.964).
